---
title: "PhySciM200 Assignment #1"
author: "Kevin Neal"
date: "April 8, 2015"
output: html_document
---

```{r init, echo=F}
require(vioplot) || {install.packages("vioplot"); library(vioplot)}
setwd("C:/Users/Kevin/Google Drive/UCLA Courses or Lab meetings etc/PhysciM200/PSM200_S15_Problem_Set_1")

### Bootstrapping function for sample medians, with 95% CI ###
# Always prints the median and 95% CI
# 10,000 bootstrap samples can be assigned to object but won't be printed to screen
bsMedian <- function(data, reps=10000) {
  bootmedian <- rep(NA, reps)
  for (i in 1:reps) {
    bootdata <- sample(data, length(data), replace=TRUE)
    bootmedian[i] <- median(bootdata)
  }
  sortmeds <- sort(bootmedian)
  dataname <- deparse(substitute(data)) # changes input data name into a string to be printed to screen
  cat("Median of ", as.character(reps), " bootstrapped ", dataname, " medians, with 95% CI: ", "\n", 
      median(bootmedian), " (", sortmeds[0.025*reps], ", ", sortmeds[0.975*reps], ")", sep="") # median of bootstrapped sample distributions/datapoints; report this with 95% CI
  invisible(bootmedian) #means the 10,000 bootstrap values will not be printed to screen but CAN be assigned to an object as if using return()
}

### MADAM
madam <- function(data) {
  med <- median(data)
  dev <- median(abs(data-med))
  return(dev)
}

### Bootstrapping function for MADAM and output
bsMadam <- function(data, reps=10000) {
  bootmadam <- rep(NA, reps)
  for (i in 1:reps) {
    bootdata <- sample(data, length(data), replace=TRUE)
    bootmadam[i] <- madam(bootdata)
  }
  sortmadam <- sort(bootmadam)
  dataname <- deparse(substitute(data)) # changes input data name into a string to be printed to screen
  cat("MADAM of ", as.character(reps), " bootstrapped ", dataname, " medians, with 95% CI: ", "\n", 
      median(bootmadam), " (", sortmadam[0.025*reps], ", ", sortmadam[0.975*reps], ")", sep="") # MADAM of bootstrapped sample distributions/datapoints; report this with 95% CI
  invisible(bootmadam) #means the 10,000 bootstrap values will not be printed to screen but CAN be assigned to an object as if using return()
}

```

### Part I: Further Reading
#### Eight ways of describing statistical methods:
1. Describe how you quantified your results and analyzed your data
2. State hypothesis you will test and the variables used to test the hypothesis
3. Consider assumptions made to allow data analysis
4. State significance level (i.e. which p-value) you chose to indicate statistical significance; justify value chosen; state if one- or two-sided test; and present actual p-value (with appropriate significant figures, e.g. with genomic analyses)
5. Do not claim groups are equivalent even if you fail to reject the null hypothesis; describe how sample size was determined in order to lessen impact of Type II error
6. A single global statistic is preferable to repeated pairwise testing; apply corrections for multiple testing and describe method used in correction
7. Choose methods that account for correlation between repeated (over time) or related/clustered observations
8. Give name, version, and source of software used to process data

### Part II: Data Analysis

```{r}
leptin <- read.csv(file="Leptin.csv", header=T) #must be in same folder as Rmd file... ugh
```

#### a. Display leptin concentrations without regard to time  
Histogram, box plot, and violin plot of 206 leptin concentrations:
```{r question a}
# how do I make captions in markdown?
lept2 <- leptin[,2] #Just "leptin" column; could also just use leptin$Leptin throughout
hist(lept2, breaks=10, density=60, freq=F, main="Histogram of leptin concentration", xlab="leptin (ng/mL)", xaxs="i", yaxs="i", tck=0.02)
curve(dnorm(x, mean=mean(lept2), sd=sd(lept2)), add=T, col="red", lwd=1) # plots normal dist based on mean and stddev of your actual data
boxplot(lept2, main="Box plot of leptin concentration", ylab="leptin (ng/mL)", tck=0.02)
vioplot(lept2, horizontal=T) #not clear to me how to title this thing or apply any normal plot args...
```


#### b. plot quantiles and determine if Gaussian

```{r question b}
qqnorm(lept2)
qqline(lept2)
```
The qqplot shows deviation from a normal/Gaussian distribution, therefore the median and 95% bootstrap confidence intervals are the better descriptors of central tendency and spread, rather than median and standard deviation.

```{r madam function}
madam = function(data) {
  med = median(data)
  dev = median(abs(data-med))
  return(dev)
}
```

```{r question b, bootstrapping median}
bsLeptin <- bsMedian(leptin$Leptin)
hist(bsLeptin, main="bootstrapped medians of leptin concentrations")

bsLeptinMadam <- bsMadam(leptin$Leptin)
hist(bsLeptinMadam, main="bootstrapped MADAMs of leptin concentration medians")

# deals = 10000
# bootmedian = rep(NA, deals)
# for (i in 1:deals) {
#   bootdata = sample(lept2, length(lept2), replace=TRUE)
#   bootmedian[i] = median(bootdata)
# }

# hist(bootmedian)
# median(bootmedian) # 9.1
# 95% Confidence intervals: sort and find 250th and 9750th values
# sortmeds <- sort(bootmedian)
# sortmeds[c(250, 9750)] # 8.8 and 9.4

# MADAM:
# samplemadam = rep(NA, deals)
# for (i in 1:deals) {
#   sampledata = sample(lept2, length(lept2), replace=TRUE)
#   samplemadam[i] = madam(sampledata)
# }
# hist(samplemadam, main="bootstrapped MADAMs of leptin concentrations")
# median(samplemadam)
# sortmadammed <- sort(samplemadam)
# sortmadammed[c(250, 9750)]

```

__Median of bootstrap-sampled medians and 95% CI: 9.1 (8.8, 9.4)__  
__Median of absolute deviation around the median (MADAM): 1.5 (1.30, 1.75)__  



```{r question c}
dayleptin <- leptin[which(leptin$Daylight==1),]
nightleptin <- leptin[which(leptin$Daylight==0),]

bsdayleptin <- bsMedian(dayleptin$Leptin)
hist(bsdayleptin, main="bootstrapped medians of daytime leptin")

bsdayleptinMadam <- bsMadam(dayleptin$Leptin)
hist(bsdayleptinMadam, main="bootstrapped MADAMs of daytime leptin")


# deals = 10000
# bootmedDay = rep(NA, deals)
# for (i in 1:deals) {
#   bootdata = sample(dayleptin$Leptin, length(dayleptin$Leptin), replace=TRUE)
#   bootmedDay[i] = median(bootdata)
# }
# #hist(bootmedDay, col="red", main="")
# median(bootmedDay) # 9.2
# # 95% Confidence intervals: sort and find 250th and 9750th values
# sortmedDay <- sort(bootmedDay)
# sortmedDay[c(250, 9750)] # 8.7, 9.7
# 
# # MADAM:
# samplemadDay = rep(NA, deals)
# for (i in 1:deals) {
#   sampledata = sample(dayleptin$Leptin, length(dayleptin$Leptin), replace=TRUE)
#   samplemadDay[i] = madam(sampledata)
# }
# #hist(samplemadDay, col="red")
# median(samplemadDay) # 1.7
# sortmadDay <- sort(samplemadDay)
# sortmadDay[c(250, 9750)] # 1.4, 1.9
```

Median of bootstrapped daytime leptin concentration: 9.2 (8.7, 9.7)
MADAM of bootstrapped daytime leptin concentration: 1.7 (1.4, 1.9)

```{r question c, night leptin}

bsnightleptin <- bsMedian(nightleptin$Leptin)
hist(bsdayleptin, main="bootstrapped medians of nighttime leptin")

bsnightleptinMadam <- bsMadam(nightleptin$Leptin)
hist(bsnightleptinMadam, main="bootstrapped MADAMs of nighttime leptin")

# 
# deals = 10000
# bootmedNi = rep(NA, deals)
# for (i in 1:deals) {
#   bootdata = sample(nightleptin$Leptin, length(nightleptin$Leptin), replace=TRUE)
#   bootmedNi[i] = median(bootdata)
# }
# hist(bootmedNi, col="blue")
# median(bootmedNi) # 9
# # 95% Confidence intervals: sort and find 250th and 9750th values
# sortmedNi <- sort(bootmedNi)
# sortmedNi[c(250, 9750)] # 8.5, 9.5
# 
# # MADAM:
# samplemadNi = rep(NA, deals)
# for (i in 1:deals) {
#   sampledata = sample(nightleptin$Leptin, length(nightleptin$Leptin), replace=TRUE)
#   samplemadNi[i] = madam(sampledata)
# }
# hist(samplemadNi, col="blue")
# median(samplemadNi) # 1.3
# sortmadNi <- sort(samplemadNi)
# sortmadNi[c(250, 9750)] # 0.8, 1.6
```

Median of 10000 sampled medians of daytime leptin concentration: 9.2 (8.7, 9.7)
MADAM of daytime leptin concentration: 1.7 (1.4, 1.9)
Median of 10000 sampled medians of nighttime leptin: 9.0 (8.5, 9.5)
MADAM of nighttime leptin concentration: 1.3 (0.8, 1.6) 


### for the time-series in questions c and d
* create a time-series vector for the data. Measurements were taken every 7 minutes, so use an interval of 420 seconds over 24 hours:
```{r}
leptin.ts <- seq(as.POSIXct("2013-10-10 08:00:00"), as.POSIXct("2013-10-11 08:00:00"), by=7*60)
leptin.ts
# add to leptin object as fourth column named "Date"
leptin[,4] <- leptin.ts
colnames(leptin)[4] <- "Date"

# re-split into day and night but now with "Date" column
dayleptin <- leptin[which(leptin$Daylight==1),]
nightleptin <- leptin[which(leptin$Daylight==0),]

plot(dayleptin$Date, dayleptin$Leptin, col="red")
points(nightleptin$Date, nightleptin$Leptin, col="navy")
```

```{r histograms of bootstrapped medians of night and day leptin concentration}
hist(bootmedNi, col="blue", xlab="median leptin concentration", main="Bootstrap resampled median Night and day leptin", tck=0.02)
hist(bootmedDay, col=rgb(1, 0, 0, 0.7), add=T)
legend("topright", legend=c("night", "day"), fill=c("blue", rgb(1, 0, 0, 0.7)))
abline(v=median(bootmedNi), lwd=2)
abline(v=median(bootmedDay), lwd=2)
box()
```

The differences in the central tendency descriptors of Daytime vs Nighttime measurements (9.2 vs 9.0), along with the difference in width of the 95% CI's (0.5 vs 0.8), suggests the two are drawn from different distributions. With two distinct clusters of data, then, each cluster ought to be analyzed separately.

#### Question d
moving average and peak-to-peak 
```{r question d}
datlen <- length(leptin$Leptin)
peaks <- rep(NA, datlen)
for (i in 2:(datlen - 1)) {
  isPeak <- leptin$Leptin[i-1] < leptin$Leptin[i] & leptin$Leptin[i] > leptin$Leptin[i+1]
  if (isPeak) # = TRUE is implicit here
    peaks[i] <- i
}
peaks <- peaks[!is.na(peaks)] # remove unused placeholders
hist(peaks)

# moving average: take avg of 3 points
movavg <- rep(NA, datlen)
for (p in 2:(datlen-1)) {
  movavg[p] <- mean( leptin$Leptin[ seq(from = p-1, to = p+1, by = 1)])
}
movavg <- movavg[!is.na(movavg)] # remove NAs
# x vector must be same length as movavg when plotting ... do same operations for the date column?
hist(movavg)
length(movavg) # length is 204



# Bootstrapping the moving average dataset
deals = 10000
bootmedMavg = rep(NA, deals)
for (i in 1:deals) {
  bootdata = sample(movavg, length(movavg), replace=TRUE)
  bootmedian[i] = median(bootdata)
}
hist(bootmedMavg)
median(bootmedMavg) # 
# 95% Confidence intervals: sort and find 250th and 9750th values
sortmedAvg <- sort(bootmedAvg)
sortmedAvg[c(250, 9750)] # 

# MADAM:
samplemadam = rep(NA, deals)
for (i in 1:deals) {
  sampledata = sample(lept2, length(lept2), replace=TRUE)
  samplemadam[i] = madam(sampledata)
}
hist(samplemadam, main="bootstrapped MADAMs of leptin concentrations")
median(samplemadam)
sortmadammed <- sort(samplemadam)
sortmadammed[c(250, 9750)]
```

?hist
# so for question d, plot histogram of bootstraps on separate day and night data